---
title: "Real-time modeling and projections of the COVID-19 epidemic in Switzerland"
author: "Christian L. Althaus, Institute of Social and Preventive Medicine, University of Bern, Switzerland"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
    html_document:
        theme: default
        toc: true
        toc_float: true
---

```{r libraries, include = FALSE}
# Load libraries and set options
library(deSolve)
library(RColorBrewer)
library(lubridate)
library(bbmle)
library(mvtnorm)

options(scipen = 10)
```

```{r data, include = FALSE}
# Load data
# Source: https://github.com/daenuprobst/covid19-cases-switzerland
reported <- read.csv("swiss_covid_epidemic.csv")
reported$date <- ymd(reported$date)
firstcase <- ymd(20200225)
lockdown <- ymd(20200317)
relaxation <- ymd(20200426)
```

```{r model, include = FALSE}
# COVID-19 transmission model
model <- function(t, x, parms) {
	with(as.list(c(parms, x)), {
	    beta <- R_0*gamma/popsize
	    if(t > (seed + control1) & t < (seed + control2)) beta <- kappa*R_0*gamma/popsize
		dS <- - beta*S*I
		dE <- beta*S*I - sigma*E
		dI <- sigma*E - gamma*I
		dP <- epsilon1*gamma*I - omega1*P
		dH1 <- omega1*P - omega2*H1
		dH2 <- (1 - epsilon2)*omega2*H1 - omega3*H2
		dU <- epsilon2*omega2*H1 - omega4*U
		dR <- (1 - epsilon1)*gamma*I + (1 - epsilon3)*omega3*H2 + (1 - epsilon4)*omega4*U
		dD <- epsilon3*omega3*H2 + epsilon4*omega4*U
		dC <- gamma*I
		der <- c(dS, dE, dI, dP, dH1, dH2, dU, dR, dD, dC)
		list(der)
	})
}

# Negative log-likelihood
nll <- function(seed, R_0, sigma, gamma, omega1, omega2, omega3, omega4, epsilon1, epsilon2, epsilon3, epsilon4, control1, control2, kappa) {
    pars <- c(seed = seed, R_0 = R_0, sigma = sigma, gamma = gamma, omega1 = omega1, omega2 = omega2, omega3 = omega3, omega4 = omega4, epsilon1 = epsilon1, epsilon2 = epsilon2, epsilon3 = epsilon3, epsilon4 = epsilon4, control1 = control1,  control2 = control2, kappa = kappa)
    pars <- trans(pars)
    times <- c(0, pars["seed"] + reported$date - min(reported$date))
    times <- c(times, max(times + 1))
    simulation <- as.data.frame(ode(inits, times, model, pars))
    simulation <- simulation[-1, ]
    ll <- sum(dpois(reported$deceased, diff(simulation$D), log = TRUE))
    return(-ll)
}

# Parameter transformations
trans <- function(pars) {
    pars["R_0"] <- exp(pars["R_0"])
    pars["seed"] <- exp(pars["seed"])
    pars["kappa"] <- plogis(pars["kappa"])
    return(pars)
}
```

```{r fit, include = FALSE, cache = TRUE}
# Fit the model to incidence of deaths
popsize <- 8.6e6
times <- 0:200			
inits <- c(S = popsize - 1,
           E = 0,
           I = 1,
           P = 0,
           H1 = 0,
           H2 = 0,
           U = 0,
           R = 0,
           D = 0,
           C = 1)
fixed <- c(sigma = 1/2.6, # Corresponds to a generation time of 5.2 days (Ganyani et al., medRxiv)
           gamma = 1/2.6, # Corresponds to a generation time of 5.2 days (Ganyani et al., medRxiv)
           omega1 = 1/5, # 5 days from onset to hospitalization
           omega2 = 1/6, # 6 days of initial hospitalization
           omega3 = 1/10, # 10 additional days of hospitalization until recovery/death
           omega4 = 1/10, # 10 additional days in ICU until recovery/death
           epsilon1 = 0.035, # Corresponds to 1.4% overall CFR
           epsilon2 = 0.3, # 30% of hospitalized cases requiring critical care in ICU
           epsilon3 = 0.35, # 35% die outside ICU ('including' deaths outside hospital)
           epsilon4 = 0.5, # 50% of those in critical care (ICU) die
           control1 = as.numeric(lockdown - firstcase),
           control2 = as.numeric(relaxation - firstcase))
free <- c(seed = log(15),
          R_0 = log(3.0),
          kappa = qlogis(0.5))
fit <- mle2(nll, start = as.list(free), fixed = as.list(fixed), method = "Nelder-Mead")
fit_ci <- confint(fit)
```

```{r bootstrap, include = FALSE, cache = TRUE}
n_sim <- 1e3
m <- coef(fit, exclude.fixed = TRUE)
sigma <- vcov(fit)
sim_coef <- data.frame(rmvnorm(n_sim, mean = m, sigma = sigma))
sim_coef$seed <- exp(sim_coef$seed)
sim_coef$R_0 <- exp(sim_coef$R_0)
sim_coef$kappa <- plogis(sim_coef$kappa)

est_R_0 <- c(trans(coef(fit))["R_0"], exp(fit_ci[2, ]))
est_kappa <- c(trans(coef(fit))["kappa"], plogis(fit_ci[3, ]))
est_R_e <- c(trans(coef(fit))["R_0"]*trans(coef(fit))["kappa"], quantile(sim_coef$R_0*sim_coef$kappa, probs = c(0.025, 0.975)))
names(est_R_e)[1] <- "R_e" 

est_R_0
est_kappa
est_R_e

mod_length <- 62 # 25 Feb 2020 to 26 Apr 2020
mod_C <- matrix(NA, nrow = mod_length, ncol = n_sim)
mod_CI <- matrix(NA, nrow = mod_length - 1, ncol = n_sim)
mod_HU <- matrix(NA, nrow = mod_length, ncol = n_sim)
mod_U <- matrix(NA, nrow = mod_length, ncol = n_sim)
mod_D <- matrix(NA, nrow = mod_length, ncol = n_sim)
mod_DI <- matrix(NA, nrow = mod_length - 1, ncol = n_sim)

for(i in 1:n_sim) {
    parms <- c(unlist(sim_coef[i, ]), fixed)
    times <- c(0, parms["seed"] + 0:(mod_length - 1))
    sim <- as.data.frame(ode(inits, times, model, parms))
    sim <- sim[-1, ]
    mod_C[, i] <- sim$C
    mod_CI[, i] <- diff(sim$C)
    mod_HU[, i] <- sim$H1 + sim$H2 + sim$U
    mod_U[, i] <- sim$U
    mod_D[, i] <- sim$D
    mod_DI[, i] <- diff(sim$D)
}

mod_C <- apply(mod_C, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
mod_CI <- apply(mod_CI, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
mod_HU <- apply(mod_HU, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
mod_U <- apply(mod_U, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
mod_D <- apply(mod_D, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
mod_DI <- apply(mod_DI, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
```

***

#### Important note: This report is work in progress and has not been peer-review. The short-term projections need to be interpreted with caution (also see Notes).

### Introduction
Switzerland reported its first case of coronavirus disease 2019 (COVID-19) on 25 Feb 2020. As of `r day(max(reported$date))` `r month(max(reported$date), label = TRUE)` `r year(max(reported$date))`, there have been `r sum(reported$confirmed)` confirmed cases and `r sum(reported$deceased)` reported deaths ([source](https://github.com/daenuprobst/covid19-cases-switzerland)). Interpreting trends in the daily numbers of reported cases can be challenging as a significant proportion of particularly mild or asymptomatic cases cannot be diagnosed and will go unreported (see [here](https://cmmid.github.io/topics/covid19/severity/global_cfr_estimates.html) for estimates of under-reporting). Here, we aim to describe the COVID-19 epidemic in Switzerland by tracking the numbers of reported deaths. To this end, we fit a dynamic transmission model to the daily number of reported deaths, estimate the reduction in transmission after the strengthening of social distancing measures on 17 Mar 2020, and project the further course of the COVID-19 epidemic in Switzerland.

### Methods
We consider a SEIR transmission model with additional compartments for hospitalization and critical care (ICU) (see [R Markdown file](swiss_covid_epidemic.Rmd) and Table). We use a maximum likelihood framework to fit the model to the [reported numbers of deaths](swiss_covid_epidemic.csv), assuming the daily numbers of deaths are Poisson distributed (see [Althaus et al.](http://dx.doi.org/10.1016/j.epidem.2015.03.001) for further details on the methods). We assume constant uncontrolled transmission until the strengthening of social distancing measures on 17 Mar 2020 and then estimate the following reduction in transmission.

**Table. Parameters of the COVID-19 transmission model.**

Parameter | Value | Source
--------- | ----- | ------
Population size of Switzerland | 8.6 million | [Federal Statistical Office](https://www.bfs.admin.ch/bfs/en/home/statistics/population.html)
Serial interval | 5.2 days | [Ganyani et al.](https://www.medrxiv.org/content/10.1101/2020.03.05.20031815v1)
Duration from onset of symptoms to hospitalization | 5 days | [Ferguson et al.](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Duration of hospitalization | 16 days | [Ferguson et al.](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Duration in critical care (ICU) | 10 days | [Ferguson et al.](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Proportion hospitalized that require critical care | 30% | [Ferguson et al.](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Proportion in critical care that will die | 50% | [Ferguson et al.](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Overall case fatality ratio | 1.4% | [Verity et al.](https://doi.org/10.1016/S1473-3099(20)30243-7) and [Wu et al.](https://doi.org/10.1038/s41591-020-0822-7)
Basic reproduction number $R_0$ | `r round(est_R_0[1], 2)` (95% CI: `r round(est_R_0[2], 2)` - `r round(est_R_0[3], 2)`) | Estimated
Reduction in transmission after 17 March 2020 | `r 1e2*round(1 - est_kappa[1], 2)`% (95% CI: `r 1e2*round(1 - est_kappa[3], 2)`%-`r 1e2*round(1 - est_kappa[2], 2)`%) | Estimated

### Results
Before 17 Mar 2020, we estimate the basic reproduction number $R_0$ of COVID-19 at `r round(est_R_0[1], 2)` (95% confidence interval, CI: `r round(est_R_0[2], 2)` - `r round(est_R_0[3], 2)`). Transmission decreased with the strengthening of social distancing measures by `r 1e2*round(1 - est_kappa[1], 2)`% (95% CI: `r 1e2*round(1 - est_kappa[3], 2)`%-`r 1e2*round(1 - est_kappa[2], 2)`%). This resulted in an effective reproduction number $R_e$ = `r round(est_R_e[1], 2)` (95% CI: `r round(est_R_e[2], 2)` - `r round(est_R_e[3], 2)`). Based on these estimates, we can project the future epidemic trajectory (Figure). The number of daily infections, hospitalized patients, patients in ICU and deaths are expected to further decline until 26 March 2020.

```{r scenarios, echo = FALSE, fig.width = 7, fig.height = 8}
# Plot best-fit model and confidence intervals
parms <- trans(coef(fit))
times <- c(0, parms["seed"] + 0:(mod_length - 1))
sim <- as.data.frame(ode(inits, times, model, parms))
sim <- sim[-1, ]
timepoints1 <- firstcase + 0:(mod_length - 1)
timepoints2 <- firstcase + 1:(mod_length - 1)

cols <- brewer.pal(4, "Set1")
t.cols <- cols
for(i in 1:length(cols)) {
    x <- col2rgb(cols[i])
    t.cols[i] <- rgb(x[1, ], x[2, ], x[3, ], alpha = 125, maxColorValue = 255)
}

par(mfrow = c(3, 2))
plot(timepoints2, diff(sim$C),
     ylim = c(1, 2e4),
     log = "y",
     ty = "l",
     col = cols[4],
     xlab = NA, ylab = "Daily number of infections", main = "Daily infections (by onset)", frame = FALSE)
polygon(x = c(timepoints2, rev(timepoints2)), y = c(mod_CI[1,], rev(mod_CI[2,])), col = t.cols[4], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)
points(reported$date - 5, reported$confirmed, pch = 22, bg = "gray")

plot(timepoints1, sim$C,
     ylim = c(0, 2e5),
     ty = "l",
     col = cols[4],
     xlab = NA, ylab = "Cumulative number of infections", main = "Total infections (by onset)", frame = FALSE)
polygon(x = c(timepoints1, rev(timepoints1)), y = c(mod_C[1,], rev(mod_C[2,])), col = t.cols[4], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)

plot(timepoints1, sim$H1 + sim$H2 + sim$U,
     ylim = c(0, 3e3),
     ty = "l",
     col = cols[3],
     xlab = NA, ylab = "Number of hospitalized patients", main = "Hospitalization", frame = FALSE)
polygon(x = c(timepoints1, rev(timepoints1)), y = c(mod_HU[1,], rev(mod_HU[2,])), col = t.cols[3], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)
points(reported$date, reported$hospitalized, pch = 22, bg = "gray")

plot(timepoints1, sim$U,
     ylim = c(0, 1e3),
     ty = "l",
     col = cols[2],
     xlab = NA, ylab = "Number of patients in ICU", main = "ICU", frame = FALSE)
polygon(x = c(timepoints1, rev(timepoints1)), y = c(mod_U[1,], rev(mod_U[2,])), col = t.cols[2], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)
points(reported$date, reported$icu, pch = 22, bg = "gray")

plot(timepoints2, diff(sim$D),
     ylim = c(0, 8e1),
     ty = "l",
     col = cols[1],
     xlab = NA, ylab = "Daily number of deaths", main = "Daily deaths", frame = FALSE)
polygon(x = c(timepoints2, rev(timepoints2)), y = c(mod_DI[1,], rev(mod_DI[2,])), col = t.cols[1], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)
points(reported$date, reported$deceased, pch = 21, bg = "white")

plot(timepoints1, sim$D,
     ylim = c(0, 2e3),
     ty = "l",
     col = cols[1],
     xlab = NA, ylab = "Cumulative number of deaths", main = "Total deaths", frame = FALSE)
polygon(x = c(timepoints1, rev(timepoints1)), y = c(mod_D[1,], rev(mod_D[2,])), col = t.cols[1], border = NA)
abline(v = lockdown, lty = 3)
abline(v = relaxation, lty = 3)
abline(v = max(reported$date), lty = 2)
points(reported$date, cumsum(reported$deceased), pch = 21, bg = "white")
```

**Figure. Projected numbers of infections, hospitalizations, patients in ICU and deaths for the COVID-19 epidemic in Switzerland.** The model was fitted to daily numbers of reported deaths (white circles). Data of confirmed cases are shifted by 5 days to account for the reporting delay. Data about hospitalizations and patients in ICU are shown for validation (gray squares). Vertical dotted and dashed lines indicate the time points of the strengthening and planned relaxation of social distancing measures (17 Mar 2020 and 26 Apr 2020, respectively) and the last data point (`r day(max(reported$date))` `r month(max(reported$date), label = TRUE)` `r year(max(reported$date))`). Note the logarithmic vertical axis for the number of daily infections.

### Notes
- 24 Apr 2020: Assuming a constant effective reprodution number $R_e$ since the strengthening of social distancing measures, the model projects a more optimistic reduction in new infections and hospitalizations than suggested by the data. This could indicate that $R_e$ slightly increased since 17 March 2020 (also see changes in [Mobility Trends Reports](https://www.apple.com/covid19/mobility) by [Apple Inc.](https://www.apple.com)).
- 14 Apr 2020: Updated model structure and parameters.
- 9 Apr 2020: Updated model structure and parameters. Data allows to estimate the reduction in transmission after the strengthening of social distancing measures.
- 24 Mar 2020: Shortening the generation time from 7.5 days ([Li et al.](https://doi.org/10.1056/NEJMoa2001316)) to 5.2 days ([Ganyani et al.](https://www.medrxiv.org/content/10.1101/2020.03.05.20031815v1)) results in a lower estimate of $R_0$, and consequently more optimistic projections about epidemic control.


### Funding
This project receives funding from the European Union's Horizon 2020 research and innovation programme - project EpiPose (No 101003688).
