---
title: "Real-time modeling and projections of the COVID-19 epidemic in Switzerland"
author: "Christian L. Althaus, Institute of Social and Preventive Medicine, University of Bern, Switzerland"
date: "23 February 2020"
output: html_document
---

```{r libraries, include = FALSE}
# Load libraries
library(deSolve)
library(RColorBrewer)
library(lubridate)
library(bbmle)
```

```{r data, include = FALSE}
# Load data
# Source: https://interaktiv.tagesanzeiger.ch/2020/wuhan-schweiz/
reported <- read.csv("swiss_covid_epidemic.csv")
reported <- reported[, 1:5]
reported <- reported[-1,]
reported[, 1] <- ymd(reported[, 1])
names(reported) <- c("date", "cum_cases", "cum_deaths", "inc_cases", "inc_deaths")
```

```{r model, include = FALSE}
# COVID-19 transmission model
model <- function(t, x, parms) {
	with(as.list(c(parms, x)), {
	    ifelse(t < (seed + control), beta <- R_0*gamma/popsize, beta <- kappa*R_0*gamma/popsize)
		dS <- - beta*S*I
		dE <- beta*S*I - sigma*E
		dI <- sigma*E - gamma*I
		dH <- epsilon1*gamma*I - omega1*H
		dV <- epsilon2*omega1*H - omega2*V
		dR <- (1 - epsilon1)*gamma*I + (1 - epsilon2)*omega1*H + (1 - epsilon3)*omega2*V
		dD <- epsilon3*omega2*V
		dC <- sigma*E
		der <- c(dS, dE, dI, dH, dV, dR, dD, dC)
		list(der)
	})
}

# Negative log-likelihood
nll <- function(seed, R_0, sigma, gamma, omega1, omega2, epsilon1, epsilon2, epsilon3, control, kappa) {
    pars <- c(seed = seed, R_0 = R_0, sigma = sigma, gamma = gamma, omega1 = omega1, omega2 = omega2, epsilon1 = epsilon1, epsilon2 = epsilon2, epsilon3 = epsilon3, control = control, kappa = kappa)
    pars <- trans(pars)
    times <- c(0, pars["seed"] + reported$date - min(reported$date))
    times <- c(times, max(times + 1))
    simulation <- as.data.frame(ode(inits, times, model, pars))
    simulation <- simulation[-1, ]
    ll <- sum(dpois(reported$inc_deaths, diff(simulation$D), log = TRUE))
    return(-ll)
}

# Parameter transformations
trans <- function(pars) {
    pars["R_0"] <- exp(pars["R_0"])
    pars["seed"] <- exp(pars["seed"])
    pars["kappa"] <- plogis(pars["kappa"])
    return(pars)
}
```

```{r fit, include = FALSE, cache = TRUE}
# Fit the model to incidence of deaths
popsize <- 8.6e6
times <- 0:200			
inits <- c(S = popsize - 1,
           E = 0,
           I = 1,
           H = 0,
           V = 0,
           R = 0,
           D = 0,
           C = 1)
fixed <- c(sigma = 1/3.75,
           gamma = 1/3.75, # Serial interval of 7.5 days (Li et al., NEJM)
           omega1 = 1/8, # 8 days hospital for mild and severe cases (Imperial College COVID-19 Response Team: Report 9)
           omega2 = 1/8, # 8 days hospital for critical cases (Imperial College COVID-19 Response Team: Report 9)
           epsilon1 = 0.05, # 5% hospitalization
           epsilon2 = 0.5, # 2.5% become critical
           epsilon3 = 0.5, # 1.25% fatal
           control = 21) # Lockdown 'light' 21 days after occurence of first case (17 March 2020)
free <- c(seed = log(30),
          R_0 = log(2.2),
          kappa = qlogis(0.1))
fit <- mle2(nll, start = as.list(free), fixed = as.list(fixed), method = "Nelder-Mead")
fit_ci <- confint(fit)
```

***

#### **Important note: This report provides preliminary results and is work in progress.**

### Introduction
Switzerland reported its first case of coronavirus disease 2019 (COVID-19) on 25 Feb 2020. As of `r day(max(reported$date))` `r month(max(reported$date), label = TRUE)` `r year(max(reported$date))`, there are `r max(reported$cum_cases)` reported cases and `r max(reported$cum_deaths)` deaths. Arguably, the reported numbers of cases do not provide a clear picture of the overall epidemic as a significant proportion of particularly mild or asymptomatic cases cannot be diagnosed and will go unreported. We argue that tracking the reported number of deaths provides a better picture of the COVID-19 epidemic in Switzerland. We fit a dynamic transmission model to the daily number of reported deaths in Switzerland, and project different future trajectories as a result of the lockdown 'light' on 17 Mar 2020.

### Methods
We consider a SEIR transmission model with additional compartments for hospitalization and critical care (ICU) (see [R Markdown file](swiss_covid_epidemic.Rmd) and Table). We use a maximum likelihood framework to fit the model to the reported numbers of deaths, assuming the daily numbers of deaths are Poisson distributed. We assume constant uncontrolled transmission until the lockdown and then vary the transmission rate relative to the epidemic spread before the lockdown.

**Table. Parameters of the COVID-19 transmission model.**

Parameter | Value | Source
--------- | ----- | ------
Population size of Switzerland | 8.6 million | [Federal Statistical Office](https://www.bfs.admin.ch/bfs/en/home/statistics/population.html)
Serial interval | 7.5 days | [Li et al.](https://doi.org/10.1056/NEJMoa2001316)
Duration of hospitalization for mild and severe cases | 8 days | [Imperial College COVID-19 Response Team: Report 9](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Additional duration of hospitalization for critical cases | 8 days | [Imperial College COVID-19 Response Team: Report 9](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Proportion hospitalized cases | 5% | Adapted from [Imperial College COVID-19 Response Team: Report 9](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Proportion critical cases | 2.5% | Adapted from [Imperial College COVID-19 Response Team: Report 9](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Overall case fatality ratio | 1.25% | Adapted from [Imperial College COVID-19 Response Team: Report 9](https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf)
Basic reproduction number $R_0$ | Estimated | -

### Results
Before the lockdown in Switzerland, we estimate the basic reproduction number $R_0$ of COVID-19 at `r round(trans(coef(fit))[["R_0"]], 2)` (95% confidence interval: `r round(exp(fit_ci[2]), 2)` - `r round(exp(fit_ci[5]), 2)`). This value of $R_0$ requires transmission to be reduced by at least `r 1e2*round(1 - 1/trans(coef(fit))[["R_0"]], 2)`% (1 - 1/$R_0$) in order to drop the effective reproduction number $R$ below the critical threshold of 1.

In order to project future epidemic trajectories, we consider the following scenarios of reduced transmission as a result of the lockdown on 17 March 2020: 100% (no reduction in transmission), 75%, 50%, 25%, 20%, 10%, and 0% (complete interruption of transmission). The model suggests that the necessary ICU capacities in Switzerland will grow beyond 1000 in early  April if the current lockdown in Switzerland does not reduce transmission by more than 75% (Figure).

```{r scenarios, echo=FALSE, fig.width=9, fig.height=6}
# Run different control scenarios
parms <- trans(coef(fit))
times <- c(0, parms["seed"])
sim <- as.data.frame(ode(inits, times, model, parms))
inits_control <- as.numeric(sim[2, 2:9])
names(inits_control) <- names(inits)
times <- 0:65

kappa_range <- c(1, 0.75, 0.5, 0.25, 0.2, 0.1, 0.0)

par(mfrow = c(2, 3))
plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(0, 5e3),
     xlab = NA, ylab = "Number of hospitalized patients", main = "Hospitalization", frame = FALSE)
cols <- brewer.pal(9, "Blues")[2:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$H + sim$V, col = cols[i])
}
legend("topleft", as.character(kappa_range), lty = rep(1, 6), col = cols, bty = "n")
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)

plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(0, 2e3),
     xlab = NA, ylab = "Number of patients in ICU", main = "ICU", frame = FALSE)
cols <- brewer.pal(9, "Purples")[3:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$V, col = cols[i])
}
legend("topleft", as.character(kappa_range), lty = rep(1, 6), col = cols, bty = "n")
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)

plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(0, 3e3),
     xlab = NA, ylab = "Cumulative number of deaths", main = "Deaths", frame = FALSE)
cols <- brewer.pal(9, "Reds")[3:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$D, col = cols[i])
}
points(reported$date, reported$cum_deaths)
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)
legend("topleft", as.character(kappa_range), lty = rep(1, 6), col = cols, bty = "n")

# Same on logarithmic scale
plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(1, 1e6), log = "y",
     xlab = NA, ylab = "Number of hospitalized patients", frame = FALSE)
cols <- brewer.pal(9, "Blues")[3:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$H + sim$V, col = cols[i])
}
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)

plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(1, 1e6), log = "y",
     xlab = NA, ylab = "Number of patients in ICU", frame = FALSE)
cols <- brewer.pal(9, "Purples")[3:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$V, col = cols[i])
}
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)

plot(ymd(20200225) + times, rep(NA, length(times)),
     ylim = c(1, 1e6), log = "y",
     xlab = NA, ylab = "Cumulative number of deaths", frame = FALSE)
cols <- brewer.pal(9, "Reds")[3:9]
for(i in 1:length(kappa_range)) {
    parms["kappa"] <- kappa_range[i]
    parms["control"] <- 21
    parms["seed"] <- 0
    sim <- as.data.frame(ode(inits_control, times, model, parms))
    sim$time <- ymd(20200225) + times
    lines(sim$time, sim$D, col = cols[i])
}
points(reported$date, reported$cum_deaths)
abline(v = ymd(20200317), lty = 3)
abline(v = max(reported$date), lty = 2)
```

**Figure. Projected numbers of hospitalizations, patients in ICU and deaths under different scenarios for the COVID-19 epidemic in Switzerland.** The different scenarios correspond to different levels of transmission relative to the epidemic spread before the lockdown. Horizontal dotted and dashed lines indicate the time points of the lockdown (17 Mar 2020) and the last data point (`r day(max(reported$date))` `r month(max(reported$date), label = TRUE)` `r year(max(reported$date))`). Epidemic trajectories are shown on a linear (top panels) and logarithmic (bottom panels) scale.
